Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-16 src/tex/ms.tex

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (HWO), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
The glint signature at planetary phase angles of 
%$\sim$50--70$^\circ$ 
$\sim$110--130$^\circ$ would be accessible to the Habitable Worlds Observatory in $\sim$14 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$42 systems, assuming a 62\,mas inner working angle (\IWA; 3$\lambda/D$ for a 6-m telescope at 600nm). Increasing the \IWA to 2$\lambda/D$, increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively. The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.     \end{abstract}

@@

Signatures of habitability like ocean glint and optical effects from condensates--- like the rainbow---should be some of the most distinct signals in an exoplanet's phase curve, particularly in polarized light.  However, depending on system architectures, some of these features can be obscured by the inner working angle of a coronagraph.

Surface liquid water is closely related to planetary habitability because it is essential for life as we know it. 
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star. The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
 However, there are many factors that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and geological activity. 
Establishing the presence of liquid surface water on an exoplanet therefore provides important context for interpreting possible biosignatures. Providing additional context to the habitability of an exoplanet is the presence of a water-cycle as evidenced by the characterization of a rainbow and other cloud optical effects.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Exoplanet Phase Curves}
An exoplanet observed in an edge-on orbit will move through all phases and thus all scattering angles, while those viewed at higher inclinations will pass through a subset of these scattering angles. If the planet is rotating, this supplies additional regular variations that can be used to place constraints on the heterogeneity of the planet and potentially map it.

The strengths of detailed phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scatter, rainbows, and glories. In polarized light an added potential benefit is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is from the specular (Fresnel) reflection of light off of smooth liquid surfaces at crescent phases. In the total flux the signal of the glint will be most prominent at small planetary phase (large scattering) angles. The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at 53 degrees  \citep[a scattering angle of 127 degrees; see e.g. ][]{2008Icar..195..927W}. It therefore varies slightly by wavelength and by sea and atmosphere composition. In general on a planet the glint will be strongest in red wavelengths where the Rayleigh contribution to both the phase curve and attenuation are minimized \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also minimized \citep{2012ApJ...752L...3C}.  The strengthened glint signature at longer wavelengths has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of surface liquid bodies, which taken in the context of the temperature and condensates may be interpreted as the presence of liquid water. The forward scattering from clouds can be confused with glint \citep{Robinson_2010} further motivating also probing the rainbow angle for context.  The cloud coverage fraction on ocean worlds can also be constrained with observations spanning the ultraviolet to near-infrared by the point phase curves at all wavelengths intersect in polarized light \citep{treesandstam2019}.  Additionally, 
 \citet{trees2022} showed that glint can also be detected in spectropolarimetry while the planet is at phase angles placing it at the wide separations more easily probed with coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

\subsection{Rainbows and Glories}
Rainbows, glories and other optical effects from condensates provide a wealth of information about an exoplanet \citep{karalidi2012rainbow, stam2008}. In our own Solar System the polarized back scatter from Titan provided evidence that the world's reflectance was due to haze rather than a surface \citep{zellner1973polarization}, on Venus the rainbow and phase curve behavior with wavelength showed the clouds were $\sim$75\% H$_2$SO$_4$ -- water mixture also constraining the droplet size distribution \citep{hansenhovenier1974}, and the glory was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light the forward scattering and backscatter from clouds can dramatically change the shape of a phase curve from Lambertian. While the glory and rainbow are technically visible in any phase curve, their strongly polarized nature coupled with the suppression of signal from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}. The rainbow angle depends on the refractive index and therefore varies with wavelength and depending on the species. The feature being far more narrow than the glint allows more exact determination of the species. The droplet size distribution also effects the shape of the rainbow and clouds of species in a solid phase can diverge from the Mie scattering that produces these features greatly, producing other optical effects particularly at narrow scattering angles \citep{bailey2007, hansentravis1974, Karalidi2012ice}. In this paper we define our optimal angle for rainbow detection as the peak of the water droplet rainbow in visible wavelengths.  However, extending observations to angles closer to the star (smaller scattering angles, i.e.\ larger phase angles) would allow the full shape of the rainbow to be characterized and ensure the characterization of rainbows from other possible HZ condensates. The glory is produced when the droplet radii much smaller than the wavelength of light and can allow further characterization as their angle, intensity, and shape also vary with the complex refractive index but primarily occur within 20 degrees scattering angle very close to full phase. %The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 habitable zone planets.
%
This assertion requires an instrument with a coronograph capable of high contrast imaging at optical to near infrared wavelengths.
%
Following this release, NASA recently announced the start of the development of the Habitable Worlds Observatory (\hwo).
%
The precursor technology recommended by the Decadal Survey also lists ``direct imaging to probe polarized ocean glint on terrestrial planets'' as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of this telescope (on-axis or off-axis, exact diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily influenced by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working angle (\IWA) of the resulting telescope will depend on these choices, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

@@

In this paper, we quantify the planetary phase angles that would be 
accessible for direct observations for hypothetical terrestrial planets 
orbiting in the habitable zones of HabWorlds target stars. 
In \S 2.1, we describe the target list, in \S 2.2 we derive expressions for the 
range of phase angles given a few different limits, in \S 2.3, we describe 
Monte Carlo simulations to marginalize over the unknown orbital inclination and 
eccentricity of planets in these systems, and in \S 2.4, we outline how the 
planet/star contrast changes as a function of the planet's orbital phase.  
We present and discuss our results in \S 3 and conclude in \S 4. 
%the   the way that inclination and coronagraph inner working angle can limit the accessible phase angles  
% We minimally explore the contrast ratio as we are primarily exploring whether these scattering angles are reachable with inner working angle and outer working angle considered.

@@

\subsection{\hwo\ and Stellar Sample}
%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo\ is still in the early stages of development, and the exact 
telescope and instrument design is as of yet undecided. 
In this work, we assume a \SI{6}{\meter} primary mirror, which is the 
currently favoured design for the observatory.
In addition, the observatory will utilise a coronagraph to suppress the 
stellar light and facilitate high-contrast observations of faint 
companying exoplanet. 
The \IWA\ of the coronagraph is currently undecided and will be driven 
by both the science requirements and technological limitations. 
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA\ of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations
corresponding to multiples of $\lambda / D$ for a \SI{6}{\meter} telescope at representative visible and near infrared wavelengths of $\lambda = \SI{600}{\nano\meter}$ and $\SI{1}{\micro\meter}$.

@@

\hwo\ is currently envisioned to observe the stars stated in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory.%
\footnote{\url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim$160 stars, the majority of which are Sun-like dwarfs; 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \SIrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \SIrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
The authors of the report determined that these are the nearby stars (maximum distance \SI{25}{\parsec}) for which exo-Earths would be the most observable for an imaging survey of habitable zones with a \SI{6}{\meter} telescope.

In what follows, we consider hypothetical Earth-size planets orbiting at the right distance from each star to receive the same flux as the Earth receives from the Sun. By construction, such planets should be observable at quadrature phase by HabWorlds.

@@

The coronagraph can obscure part of the planet's orbit which, as demonstrated in \cref{fig:annotated-orbit}, can prevent the planet from being observed at high and low scattering angles. 
%
The range of scattering phase angles that can be observed will depend on the \IWA\ of the coronagraph as well as the shape and orientation of the exoplanet's orbit.

\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        An illustration of an edge-on circular orbit showing that the coronagraph can obscure the exoplanet at high and low scattering phase angles. This can prevent observations of scattering phenomena which could be used to indicate the habitability and atmospheric composition of the exoplanet. 
        %\timmy{Maybe mention that this is a simplified illustration with a circular orbit, plus we are basically looking at an edge-on system?}
    }
    \label{fig:annotated-orbit}
\end{figure}

@@

The on-sky projected planet distance, $r_\mathrm{proj}$ (in \si{\au}) is shown in \cref{fig:scattering-angle} assuming a non-eccentric circular orbit with a semi-major $a$ (in \si{\au}) at a distance $d_*$ (in \si{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ in this set up follows the equation,
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}
 %
\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The on-sky projected planet distance, $r_\mathrm{proj}$, of a planet with a semi-major $a$ at a distance $d_*$. The on-sky angular separation as viewed by the observer is $\delta$. The scattering phase angle, $\varphi$, is the angle between the line of sight to the star and the line connecting the star and planet.  
    }
    \label{fig:scattering-angle}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics{figures/scatterplot.pdf}
    \caption{
        Scatter plot for the target sample, showing stellar effective temperature and stellar distance. 
        The size of the points represents the angular separation of the star and planet in milliarcseconds as presented in the target list. 
        The colour of the points shows $\Delta \varphi$ assuming a circular edge-on orbit at a semi-major axis corresponding to an Earth-like instellation for an IWA of 62 mas. Additionally, the colour bar indicates the atmospheric phenomenon that can be detected where phenomena from the bottom of the colour bar up to the indicated colour are detectable for that system.
        Thus, dark blue points are systems which have the most key features, as systems in which the angles required to see the rainbow are probed will also have the angles required to see the Rayleigh scattering probed.
    }
    \label{fig:scatterplot}
    \script{create-scatterplot.py}
\end{figure*}

@@

(1)~If the inclination is relatively face-on, then the coronagraph mask does not obscure any part of the sky-projected orbit, and the maximum scattering phase angle coverage depends solely on the disk inclination ;

(2)~When parts of the orbit  are hidden by the focal plane mask, the scattering phase angle coverage only depends on the \IWA\ and orbital semi-major axis.

\begin{figure}%[htb]
   \centering
   \includegraphics[width=0.99\columnwidth]{figures/orb-grid.pdf}
   \caption{
        The scattering phase angles accessible to a direct imaging mission can be limited by orbital inclination or by the coronagraphic inner working angle. 
        If the orbital inclination is close to face-on (top panels) then the scattering phase angles probed throughout an orbit are limited by inclination; for circular orbits accessible scattering phases are $\phi \in 90^\circ \pm i$. 
        For orbital inclinations close to edge-on (bottom panels) the planet goes through all phases, but the scattering phase angles near 0$^\circ$ and 180$^\circ$ are not accessible due to obscuration by the coronagraph mask; for circular orbits, the accessible scattering phases are $\phi \in 90^\circ \pm \arcsin({\rm IWA}\cdot d_*/a)$.
    }
    \label{fig:orb-grid}
    \script{plot-orb-grid.py}
\end{figure}

>>>>

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (HWO), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
%
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
%
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
%
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
%
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
%
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
%
The glint signature at planetary phase angles of 
%$\sim$50--70$^\circ$ 
$\sim$110--130$^\circ$ would be accessible to the Habitable Worlds Observatory in $\sim$14 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$42 systems, assuming a 62\,mas inner working angle (\IWA; 3$\lambda/D$ for a 6-m telescope at 600nm).
%
Increasing the \IWA to 2$\lambda/D$, increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.     \end{abstract}

@@

Signatures of habitability, from ocean glint to optical effects from condensates --- like the rainbow---should be some of the most distinct signals in an exoplanet's phase curve, particularly in polarized light.
%
However, depending on system architectures, some of these features can be obscured by the inner working angle of a coronagraph.

Surface liquid water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which a rocky planet (or moon) may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet therefore provides important context for interpreting possible biosignatures.
%
Providing additional context to the habitability of an exoplanet is the presence of a water-cycle as evidenced by the characterization of a rainbow and other cloud optical effects.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Exoplanet Phase Curves}

An exoplanet observed in an edge-on orbit will move through all phases and thus all scattering angles, while those viewed at higher inclinations will pass through a subset of these scattering angles.
%
If the planet is rotating, this supplies additional regular variations that can be used to place constraints on the heterogeneity of the planet and potentially map it.

The strengths of detailed phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scatter, rainbows, and glories.
%
In polarized light an added potential benefit is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is from the specular (Fresnel) reflection of light off of smooth liquid surfaces at crescent phases.
%
In the total flux the signal of the glint will be most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at 53 degrees  \citep[a scattering angle of 127 degrees; see e.g. ][]{2008Icar..195..927W}.
%
It therefore varies slightly by wavelength and by sea and atmosphere composition.
%
In general on a planet the glint will be strongest in red wavelengths where the Rayleigh contribution to both the phase curve and attenuation are minimized \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also minimized \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of surface liquid bodies, which taken in the context of the temperature and condensates may be interpreted as the presence of liquid water.
%
The forward scattering from clouds can be confused with glint \citep{Robinson_2010} further motivating also probing the rainbow angle for context.
%
The cloud coverage fraction on ocean worlds can also be constrained with observations spanning the ultraviolet to near-infrared by the point phase curves at all wavelengths intersect in polarized light \citep{treesandstam2019}.
%
Additionally, \citet{trees2022} showed that glint can also be detected in spectropolarimetry while the planet is at phase angles placing it at the wide separations more easily probed with coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

\subsection{Rainbows and Glories}

Rainbows, glories and other optical effects from condensates provide a wealth of information about an exoplanet \citep{karalidi2012rainbow, stam2008}.
%
In our own Solar System the polarized back scatter from Titan provided evidence that the world's reflectance was due to haze rather than a surface \citep{zellner1973polarization}, on Venus the rainbow and phase curve behavior with wavelength showed the clouds were $\sim$75\% H$_2$SO$_4$ -- water mixture also constraining the droplet size distribution \citep{hansenhovenier1974}, and the glory was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light the forward scattering and backscatter from clouds can dramatically change the shape of a phase curve from Lambertian.
%
While the glory and rainbow are technically visible in any phase curve, their strongly polarized nature coupled with the suppression of signal from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index and therefore varies with wavelength and depending on the species.
%
The feature being far more narrow than the glint allows more exact determination of the species.
%
The droplet size distribution also effects the shape of the rainbow and clouds of species in a solid phase can diverge from the Mie scattering that produces these features greatly, producing other optical effects particularly at narrow scattering angles \citep{bailey2007, hansentravis1974, Karalidi2012ice}.
%
In this paper we define our optimal angle for rainbow detection as the peak of the water droplet rainbow in visible wavelengths.
%
However, extending observations to angles closer to the star (smaller scattering angles, i.e.\ larger phase angles) would allow the full shape of the rainbow to be characterized and ensure the characterization of rainbows from other possible HZ condensates.
%
The glory is produced when the droplet radii much smaller than the wavelength of light and can allow further characterization as their angle, intensity, and shape also vary with the complex refractive index but primarily occur within 20 degrees scattering angle very close to full phase. 
%
%The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 habitable zone planets.
%
This assertion requires an instrument with a coronagraph capable of high contrast imaging at optical to near infrared wavelengths.
%
Following this release, NASA recently announced the start of the development of the Habitable Worlds Observatory (\hwo).
%
The precursor technology recommended by the Decadal Survey also lists ``direct imaging to probe polarized ocean glint on terrestrial planets'' as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of this telescope (on-axis or off-axis, exact diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily influenced by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working angle (\IWA) of the resulting telescope will depend on these choices, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

@@

In this paper, we quantify the planetary phase angles that would be accessible for direct observations for hypothetical terrestrial planets orbiting in the habitable zones of HabWorlds target stars.
%
In \S 2.1, we describe the target list, in \S 2.2 we derive expressions for the range of phase angles given a few different limits, in \S 2.3, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets in these systems, and in \S 2.4, we outline how the planet/star contrast changes as a function of the planet's orbital phase.  
%
We present and discuss our results in \S 3 and conclude in \S 4. 
%the   the way that inclination and coronagraph inner working angle can limit the accessible phase angles  
% We minimally explore the contrast ratio as we are primarily exploring whether these scattering angles are reachable with inner working angle and outer working angle considered.

@@

\subsection{\hwo\ and Stellar Sample}
%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo\ is still in the early stages of development, and the exact telescope and instrument design is as of yet undecided. 
%
In this work, we assume a \SI{6}{\meter} primary mirror, which is the currently favoured design for the observatory.
%
The observatory will utilise a coronagraph to suppress the stellar light and facilitate high-contrast observations of faint accompanying exoplanet. 
%
The \IWA{} of the coronagraph is currently undecided and will be driven by both the science requirements and technological limitations. 
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA\ of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a \SI{6}{\meter} telescope at representative visible and near infrared wavelengths of $\lambda = \SI{600}{\nano\meter}$ and $\SI{1}{\micro\meter}$.

@@

\hwo\ is currently envisioned to observe the stars stated in the NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory.
%
\footnote{\url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list comprises $\sim$160 stars, the majority of which are Sun-like dwarfs; 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude of $R_c = 31$ and a contrast floor of \num{2.5e-11}. 
%with an occurrence rate of rocky planets in the optimistic habitable zone to be $\eta_{\oplus}$ = 0.24 \citep{decadal}. 
The corresponding adopted habitable zone limits are a semi-major axis of \SIrange{0.95}{1.67}{\au} for a solar twin, planet sizes between \SIrange{0.8}{1.4}{} Earth radii, and for non-solar stars, scale as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
The authors of the report determined that these are the nearby stars (maximum distance \SI{25}{\parsec}) for which exo-Earths would be the most observable for an imaging survey of habitable zones with a \SI{6}{\meter} telescope.

In what follows, we consider hypothetical Earth-size planets orbiting at the right distance from each star to receive the same flux as the Earth receives from the Sun.
%
By construction, such planets should be observable at quadrature phase by HabWorlds.

@@

The coronagraph can obscure part of the planet's orbit which, as demonstrated in \cref{fig:annotated-orbit}, can prevent the planet from being observed at high and low scattering angles. 
%
The range of scattering phase angles that can be observed will depend on the \IWA{} of the coronagraph as well as the shape and orientation of the exoplanet's orbit.

\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        An illustration of an edge-on circular orbit showing that the coronagraph can obscure the exoplanet at high and low scattering phase angles.
        %
        This can prevent observations of scattering phenomena which could be used to indicate the habitability and atmospheric composition of the exoplanet. 
        %\timmy{Maybe mention that this is a simplified illustration with a circular orbit, plus we are basically looking at an edge-on system?}
    }
    \label{fig:annotated-orbit}
\end{figure}

@@

The on-sky projected planet distance, $r_\mathrm{proj}$ (in \si{\au}) is shown in \cref{fig:scattering-angle} assuming a non-eccentric circular orbit with a semi-major $a$ (in \si{\au}) at a distance $d_*$ (in \si{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ in this set up follows the equation,
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}
 %
\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The on-sky projected planet distance, $r_\mathrm{proj}$, of a planet with a semi-major $a$ at a distance $d_*$.
        %
        The on-sky angular separation as viewed by the observer is $\delta$.
        %
        The scattering phase angle, $\varphi$, is the angle between the line of sight to the star and the line connecting the star and planet.  
    }
    \label{fig:scattering-angle}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics{figures/scatterplot.pdf}
    \caption{
        Scatter plot for the target sample, showing stellar effective temperature and stellar distance.
        %
        The size of the points represents the angular separation of the star and planet in milliarcseconds as presented in the target list.
        %
        The colour of the points shows $\Delta \varphi$ assuming a circular edge-on orbit at a semi-major axis corresponding to an Earth-like instellation for an IWA of 62 mas.
        %
        Additionally, the colour bar indicates the atmospheric phenomenon that can be detected where phenomena from the bottom of the colour bar up to the indicated colour are detectable for that system.
        %
        Thus, dark blue points are systems which have the most key features, as systems in which the angles required to see the rainbow are probed will also have the angles required to see the Rayleigh scattering probed.
    }
    \label{fig:scatterplot}
    \script{create-scatterplot.py}
\end{figure*}

@@

\begin{enumerate}
    \item If the inclination is relatively face-on, then the coronagraph mask does not obscure any part of the sky-projected orbit, and the maximum scattering phase angle coverage depends solely on the disk inclination ;
    \item When parts of the orbit  are hidden by the focal plane mask, the scattering phase angle coverage only depends on the \IWA{} and orbital semi-major axis. 
\end{enumerate}

\begin{figure}%[htb]
   \centering
   \includegraphics[width=0.99\columnwidth]{figures/orb-grid.pdf}
   \caption{
        The scattering phase angles accessible to a direct imaging mission can be limited by orbital inclination or by the coronagraphic inner working angle.
        %
        If the orbital inclination is close to face-on (top panels) then the scattering phase angles probed throughout an orbit are limited by inclination; for circular orbits accessible scattering phases are $\phi \in 90^\circ \pm i$. 
        %
        For orbital inclinations close to edge-on (bottom panels) the planet goes through all phases, but the scattering phase angles near 0$^\circ$ and 180$^\circ$ are not accessible due to obscuration by the coronagraph mask; for circular orbits, the accessible scattering phases are $\phi \in 90^\circ \pm \arcsin({\rm IWA}\cdot d_*/a)$.
    }
    \label{fig:orb-grid}
    \script{plot-orb-grid.py}
\end{figure}
